Practical Performance Evaluation of Entity Resolution Algorithms: Lessons Learned at PatentsView.org

### 2022-07-21 paper/main.tex
Updates from Overleaf

@@

\author[1,2]{Olivier Binette}
\author[1]{Youngsoo Baek}
\author[2]{Sokhna A York}
\author[2]{Emma Hickerson}
\author[2]{Sarvo Madhavan}
\author[2]{Christina Jones}
\affil[1]{Department of Statistical Science, Duke University}
\affil[2]{American Institutes for Research}

@@

\begin{abstract}
This paper introduces a novel evaluation methodology for entity resolution algorithms. It is motivated by PatentsView.org, a \ob{U.S. Patents and Trademarks Office patent data exploration tool} which disambiguates patent inventors using an entity resolution algorithm. We provide a data collection methodology and tailored performance estimators which account for sampling biases. Our approach is simple, practical and principled -- key characteristics which allow us to paint the first representative picture of PatentsView's disambiguation performance. This is used to inform PatentsView's users of the reliability of the data and to allow the comparison of competing disambiguation algorithms.
\end{abstract}

@@

First, entity resolution problems \ob{do not scale linearly}. While it may be easy to disambiguate a small dataset, the opportunity for errors grows \textit{quadratically} in the number of records. As such, we may observe good performance of an algorithm on a small benchmark dataset, while the true performance of on the entire data may be something else entirely.\footnote{This particular effect of dataset size in entity resolution is explored in \cite{Draisbach2013} in the context of choosing similarity thresholds.} This is a problem that PatentsView.org currently faces. Despite encouraging performance evaluation metrics on benchmark datasets, with nearly perfect precision and recall reported in the latest methodological report \citep{Monath2021}, the data science team at AIR observes lower real-world accuracy. \ob{This phenomenon is illustrated in example \ref{first_example} below.}

A second problem is large class imbalance in entity resolution \citep{Marchant2017}. \ob{Viewing entity resolution as a classification problem, the task is to classify record pairs as being a match or non-match. However,} among all pairs of records, only a small fraction (usually much less than a fraction of a percent) refer to the same entity. The vast majority of record pairs are not a match. This makes it difficult to evaluate performance through random sampling of record pairs.

@@

\begin{example}[Bias of precision computed on benchmark datasets]\label{first_example}
    To exemplify the problem with the trivial use of performance evaluation metrics on benchmark datasets, we have carried out a simple experiment which is described in detail in appendix \ref{appendix:example}. In short, we considered evaluating a disambiguation algorithm by sampling ground truth clusters and computing pairwise precision\footnote{Pairwise precision is the probability that a pair of record predicted to be a match is indeed a match. See section \ref{sec:proposed_estimators} for a formal definition} on this set of sampled clusters. This is analogous to the way that many real-world benchmark data sets are obtained and used. In this experiment, we known that the disambiguation algorithm has a precision of $52\%$ for the entire dataset.

@@

    In panel \textbf{A} of figure \ref{fig:precision_problem} below, we see the distribution of precision estimates versus the true precision of $52\%$ shown as a dotted vertical line. Precision estimates are most of the time very close to $100\%$ and always higher than $80\%$, despite the truth being a precision of only $52\%$. In contrast, panel \textbf{B} shows the distribution of our proposed precision estimator which is nearly unbiased. Both precision estimators rely on exactly the same data. They only differ in how they account for the underlying sampling process and the extrapolation from small benchmark datasets to full data.

@@

\ob{While our work focuses on estimating pairwise precision and recall, the general approach also applies to any cluster based performance evaluation metrics, such as those described in \cite{Michelson2009}.}

@@

\ob{In contrast to the above, our approach to estimation is more practical than pairwise sampling and applies to any black-box disambiguation algorithms such as those used at PatentsView.}

@@

\ob{Furthermore, our approach is novel. To our knowledge, we are the first to propose unbiased performance estimators based on cluster and block samples. Past work either ignored biases when computing precision and recall from benchmark datasets \citep{frisoli2018exploring, Monath2021, han2019disambiguating}, did not provide estimates for one of precision or recall \citep{mcveigh2019scaling}, or provided solutions that are tailored to very specific record linkage models \citep{Belin1995}. We provide the first general solution to entity resolution evaluation which does not rely on sampling record pairs and which applies to any disambiguation algorithm.}

@@

The rest of the paper is organized as follows. In section \ref{sec:data_methods}, \ob{we describe: motivating benchmark datasets, our data collection process for evaluation, the proposed estimators, and our simulation study used to validate the performance of our estimators.} Section \ref{sec:results} then shows the results: our performance estimates and results from the simulation study. Finally, section \ref{sec:discussion} summarizes the paper and discusses future research directions.

@@

In this section, \ob{we introduce: motivating datasets used at PatentsView, our hand-disambiguation methodology, our proposed performance metric estimators, and our simulation framework used to compare estimators.} \ob{Note that we focus on \textit{inventor} disambiguation throughout, rather than the related problems of assignee and location disambiguation.}

@@

In addition to the above benchmark datasets, \ob{we have carried out a data collection experiment. This was motivated by the evaluation of the current PatentsView inventor disambiguation using the estimators proposed in section \ref{sec:proposed_estimators}.}

@@

Two AIR staff were tasked with recovering inventors' patents given sampled inventor mentions. To do this, they were assisted by tools available on PatentsView.org. First, given a sampled inventor, the associated predicted cluster was reviewed and wrongly assigned patents were removed. Next, PatentsView's search tools were used to find additional mentions of similarly named inventors. These were reviewed and added to the predicted cluster if appropriate. The two AIR staff had an initial training session, followed by a test run on 10 inventors, before carrying out the rest of the data collection. They worked independently so that two resulting datasets were obtained for the same inventor mentions. In section \ref{sec:results}, these are referred to as the \textbf{Staff 1} and \textbf{Staff 2} datasets.

Note that our data collection methodology is biased towards PantentsView's current disambiguation. \ob{Indeed, we do not expect the staff to have found all errors or all missing inventor mentions from predicted clusters. The staff used their best judgment, supported by a thorough search, to resolve inventor mentions.} In cases where no errors were found, the current disambiguation was assumed to be correct. Performance estimates based on this data might therefore be slightly optimistic. \ob{This should be acknowledged when reporting performance estimates to PatentsView.org users. Otherwise, for the purpose of improving the current disambiguation algorithm, the slightly biased data is still appropriate to use. It represents the most visible errors in the current disambiguation rather than the totality of them.}

@@

All of the expressions for precision and recall in lemmas \ref{lemma:record_sampling_representation} -- \ref{lemma:block_sampling_representation} are either population means or ratios of population means. As such, they can be estimated using sample means and ratios of sample means. For readability, we present here a generic formula for an approximately unbiased estimator of the ratio of means, and then specify the needed quantities for each representation below. The estimator applies a first order bias correction to the ratio of sample means, based on a Taylor approximation. Approximate confidence intervals can be computed based on the variance estimator of the ratio of sample means by Taylor approximation, assuming the corrected estimator has small bias \citep{fuller2011sampling}. The generic formula for estimating the ratio of $T$-sized ``population'' (of records/clusters/blocks) means of the form
\begin{align}
    E=\frac{1/T\sum_{i=1}^{T} B_i}{1/T\sum_{i=1}^T A_i},
\end{align}
assuming we have sampled $n$ elements (records/clusters/blocks), is
\begin{align}\label{eq:bias_adjustment}
    \widehat{E} = \frac{\bar{B}_n}{\bar{A}_n} \left\{ 1 + \frac{\theta_{n,T}}{n(n-1)} \sum_{s=1}^n \frac{A_s}{\bar{A}_n} \left(\frac{B_s}{\bar{B}_n} - \frac{A_s}{\bar{A}_n} \right) \right\}, \; \bar{A}_n = \frac{1}{n} \sum_{s=1}^{n} A_s, \; \bar{B}_n = \frac{1}{n} \sum_{s=1}^{n} B_s
\end{align}
We note that an additional symbol $\theta_{n,T}$ is introduced for possible finite population correction when relatively large number of elements are sampled without replacement (see below). Classical adjustment is set to $\theta_{n,T} = (1 - \frac{n-1}{T-1})$ \citep{Cochran1977}. For practical purposes when $T$ is large, often $\theta_{n,T}=1$ will suffice. In fact, knowledge of $T$ is not needed at all as long as it is large enough relative to $n$, which is useful because the total number of true clusters/blocks is not known in advance. Confidence interval can be computed based on the variance estimate of the above, which is
\begin{align} \label{eq:estimated_variance}
    \widehat{V}(\widehat{E}) = \left( \frac{\bar{B}_n}{\bar{A}_n} \right)^2 \frac{\theta_{n,T}}{n(n-1)} \sum_{s=1}^n \left( \frac{A_s}{\bar{A}_n} - \frac{B_s}{\bar{B}_n}\right)^2.
\end{align}
The specific values for $A_s$, $B_s$, and $T$ in each of our representation is described in appendix \ref{sec:appendix}.

@@

\begin{table}[!h]
    \label{tab:results}
    \centering
    \caption{\centering Estimated pairwise precision and recall (with estimated standard deviation) from our four benchmark datasets.}
    \begin{tabular}{lccc}
    \toprule
    \textbf{dataset} & \textbf{est. precision ($\hat \sigma$)} & \textbf{est. recall ($\hat \sigma$)} & \textbf{scope} \\
    \midrule
    Staff 1 & 86\% (5.6\%) & 94\% (2.1\%)  & 1976 -- Dec. 31, 2022 (U.S. granted)\\
    Staff 2 & 88\% (5.6\%) & 95\% (2.0\%) & 1976 -- Dec. 31, 2022 (U.S. granted)\\
    Israeli Benchmark & 79\% (NA) & 94\% (NA) & 1976 -- 1999 (U.S. granted)\\
    \cite{Li2014}'s Benchmark & 91\% (2.7\%) & 91\% (5.0\%) & 1976 -- 2010 (U.S. granted)\\
    \bottomrule
    \end{tabular}
\end{table}

@@

There are two main resulting products for this work. The first is to provide PatentsView data users with an appropriate understanding of PatentsView data quality. Our performance estimates indicate that, despite an overall accurate disambiguation, there is room for improvement and the disambiguation algorithm estimates more links then there truly are. The second product is the set of tools needed for methodological research and model comparison. Given our evaluation methodology, we can now reliably compare algorithms and decide with confidence on changes which will affect users.

\ob{One important topic for future work is the quantification of uncertainty associated with errors in the hand-disambiguation process. This is challenging problem given the lack of validation information available. Surveying inventors to validate the hand-disambiguation process would be one approach to explore for this. Sensitivity analyses involving potential errors in the hand-disambiguation process could also be informative. We refer the reader to \citep{bailey2017well} for a state-of-the-art hand-labeling study which evaluated human review accuracy.}

\ob{Another important topic for future work is the development of estimators for additional performance metrics. As we have seen in the simulation study, the generically-derived estimators (e.g., from lemma \ref{lemma:cluster_sampling_representation}) do not perform as well as estimators derived using specific properties of pairwise precision and recall (lemma \ref{lemma:block_sampling_representation}). As such, care should be taken to obtain efficient estimators for every metric of interest.}

@@

Olivier Binette led the project and wrote the bulk of the manuscript. Youngsoo Baek provided bias adjustment and uncertainty quantification for ratio estimators. Sokhna A York and Emma Hickerson carried out data collection by manually reviewing inventor clusters. Sarvo Madhavan was a technical advisor and contributed to code. Christina Jones was an advisor and project manager. All authors provided input on the manuscript.

>>>>

\usepackage{subcaption}
\DeclareCaptionFormat{custom}
{%
    \textbf{#1#2}\textit{#3}
}
\captionsetup{format=custom}

@@

\author[1,2]{Olivier Binette}
\author[2]{Sokhna A York}
\author[2]{Emma Hickerson}
\author[1]{Youngsoo Baek}
\author[2]{Sarvo Madhavan}
\author[2]{Christina Jones}
\affil[1]{Duke University}
\affil[2]{American Institutes for Research}

@@

\begin{abstract}
This paper introduces a novel evaluation methodology for entity resolution algorithms. It is motivated by PatentsView.org, a {U.S. Patents and Trademarks Office patent data exploration tool} which disambiguates patent inventors using an entity resolution algorithm. We provide a data collection methodology and tailored performance estimators which account for sampling biases. Our approach is simple, practical and principled -- key characteristics which allow us to paint the first representative picture of PatentsView's disambiguation performance. This is used to inform PatentsView's users of the reliability of the data and to allow the comparison of competing disambiguation algorithms.
\end{abstract}

@@

First, entity resolution problems {do not scale linearly}. While it may be easy to disambiguate a small dataset, the opportunity for errors grows \textit{quadratically} in the number of records. As such, we may observe good performance of an algorithm on a small benchmark dataset, while the true performance of on the entire data may be something else entirely.\footnote{This particular effect of dataset size in entity resolution is explored in \cite{Draisbach2013} in the context of choosing similarity thresholds.} This is a problem that PatentsView.org currently faces. Despite encouraging performance evaluation metrics on benchmark datasets, with nearly perfect precision and recall reported in the latest methodological report \citep{Monath2021}, the data science team at AIR observes lower real-world accuracy. {This phenomenon is illustrated in example \ref{first_example} below.}

A second problem is large class imbalance in entity resolution \citep{Marchant2017}. {Viewing entity resolution as a classification problem, the task is to classify record pairs as being a match or non-match. However,} among all pairs of records, only a small fraction (usually much less than a fraction of a percent) refer to the same entity. The vast majority of record pairs are not a match. This makes it difficult to evaluate performance through random sampling of record pairs.

@@

\begin{example}[Bias of precision computed on benchmark datasets]\label{first_example}
    To exemplify the problem with the trivial use of performance evaluation metrics on benchmark datasets, we have carried out a simple experiment which is described in detail in appendix \ref{appendix:example}. In short, we considered evaluating a disambiguation algorithm by sampling ground truth clusters and computing pairwise precision\footnote{Pairwise precision is the probability that a pair of record predicted to be a match is indeed a match. See section \ref{sec:proposed_estimators} for a formal definition} on this set of sampled clusters. This is analogous to the way that many real-world benchmark data sets are obtained and used. In this experiment, we know that the disambiguation algorithm has a precision of $52\%$ for the entire dataset.

@@

    In panel \textbf{A} of figure \ref{fig:precision_problem}, we see the distribution of precision estimates versus the true precision of $52\%$ shown as a dotted vertical line. Precision estimates are most of the time very close to $100\%$ and always higher than $80\%$, despite the truth being a precision of only $52\%$. In contrast, panel \textbf{B} shows the distribution of our proposed precision estimator which is nearly unbiased. Both precision estimators rely on exactly the same data. They only differ in how they account for the underlying sampling process and the extrapolation from small benchmark datasets to full data.

@@

{While our work focuses on estimating pairwise precision and recall, the general approach also applies to any cluster based performance evaluation metrics, such as those described in \cite{Michelson2009}.}

@@

{In contrast to the above, our approach to estimation is more practical than pairwise sampling and applies to any black-box disambiguation algorithms such as those used at PatentsView.}

@@

{Furthermore, our approach is novel. To our knowledge, we are the first to propose unbiased performance estimators based on cluster and block samples. Past work either ignored biases when computing precision and recall from benchmark datasets \citep{frisoli2018exploring, Monath2021, han2019disambiguating}, did not provide estimates for one of precision or recall \citep{mcveigh2019scaling}, or provided solutions that are tailored to very specific record linkage models \citep{Belin1995}. We provide the first general solution to entity resolution evaluation which does not rely on sampling record pairs and which applies to any disambiguation algorithm.}

@@

The rest of the paper is organized as follows. In section \ref{sec:data_methods}, {we describe: motivating benchmark datasets, our data collection process for evaluation, the proposed estimators, and our simulation study used to validate the performance of our estimators.} Section \ref{sec:results} then shows the results: our performance estimates and results from the simulation study. Finally, section \ref{sec:discussion} summarizes the paper and discusses future research directions.

@@

In this section, {we introduce: motivating datasets used at PatentsView, our hand-disambiguation methodology, our proposed performance metric estimators, and our simulation framework used to compare estimators.} {Note that we focus on \textit{inventor} disambiguation throughout, rather than the related problems of assignee and location disambiguation.}

@@

In addition to the above benchmark datasets, {we have carried out a data collection experiment. This was motivated by the evaluation of the current PatentsView inventor disambiguation using the estimators proposed in section \ref{sec:proposed_estimators}.}

@@

Two AIR staff were tasked with recovering inventors' patents given sampled inventor mentions. To do this, they were assisted by tools available on PatentsView.org. First, given a sampled inventor, the associated predicted cluster was reviewed \ob{and, if any, wrongly} assigned patents were removed. Next, PatentsView's search tools were used to find additional mentions of similarly named inventors. These were reviewed \ob{and inventor mentions were added} to the predicted cluster if appropriate. The two AIR staff had an initial training session, followed by a test run on 10 inventors, before carrying out the rest of the data collection. They worked independently so that two resulting datasets were obtained for the same inventor mentions. In section \ref{sec:results}, these are referred to as the \textbf{Staff 1} and \textbf{Staff 2} datasets.

Note that our data collection methodology is biased towards PantentsView's current disambiguation. {Indeed, we do not expect the staff to have found all errors or all missing inventor mentions from predicted clusters. The staff used their best judgment, supported by a thorough search, to resolve inventor mentions.} In cases where no errors were found, the current disambiguation was assumed to be correct. Performance estimates based on this data might therefore be slightly optimistic. {This should be acknowledged when reporting performance estimates to PatentsView.org users. Otherwise, for the purpose of improving the current disambiguation \ob{algorithm, this data} is still appropriate to use. It represents the most visible errors in the current disambiguation rather than the totality of them.}

@@

All of the expressions for precision and recall in lemmas \ref{lemma:record_sampling_representation} -- \ref{lemma:block_sampling_representation} are either population means or ratios of population means. As such, they can be estimated using sample means and ratios of sample means. For readability, we present here a generic formula for an approximately unbiased estimator of the ratio of means, and then specify the needed quantities for each representation below. The estimator applies a first order bias correction to the ratio of sample means, based on a Taylor approximation. Approximate confidence intervals can be computed based on the variance estimator of the ratio of sample means by Taylor approximation, assuming the corrected estimator has small bias \citep{fuller2011sampling}. The generic formula for estimating the ratio of $T$-sized ``population'' (of records/clusters/blocks) means of the form
\begin{align}
    E=\frac{1/T\sum_{i=1}^{T} B_i}{1/T\sum_{i=1}^T A_i},
\end{align}
assuming we have sampled $n$ elements (records/clusters/blocks), is
\begin{align}\label{eq:bias_adjustment}
    \widehat{E} = \frac{\bar{B}_n}{\bar{A}_n} \left\{ 1 + \frac{\theta_{n,T}}{n(n-1)} \sum_{s=1}^n \frac{A_s}{\bar{A}_n} \left(\frac{B_s}{\bar{B}_n} - \frac{A_s}{\bar{A}_n} \right) \right\}, \; \bar{A}_n = \frac{1}{n} \sum_{s=1}^{n} A_s, \; \bar{B}_n = \frac{1}{n} \sum_{s=1}^{n} B_s
\end{align}
We note that an additional symbol $\theta_{n,T}$ is introduced for possible finite population correction when relatively large number of elements are sampled without replacement (see below). Classical adjustment is set to $\theta_{n,T} = (1 - \frac{n-1}{T-1})$ \citep{Cochran1977}. For practical purposes when $T$ is large, often $\theta_{n,T}=1$ will suffice. In fact, knowledge of $T$ is not needed at all as long as it is large enough relative to $n$, which is useful because the total number of true clusters/blocks is not known in advance. Confidence interval can be computed based on the variance estimate of the above, which is
\begin{align} \label{eq:estimated_variance}
    \widehat{V}(\widehat{E}) = \left( \frac{\bar{B}_n}{\bar{A}_n} \right)^2 \frac{\theta_{n,T}}{n(n-1)} \sum_{s=1}^n \left( \frac{A_s}{\bar{A}_n} - \frac{B_s}{\bar{B}_n}\right)^2.
\end{align}
The specific values for $A_s$, $B_s$, and $T$ in each of our representation are described in appendix \ref{sec:appendix}.

@@

\begin{table}
    \label{tab:results}
    \centering
    \caption{\centering Estimated pairwise precision and recall (with estimated standard deviation) from our four benchmark datasets.}
    \begin{tabular}{lccc}
    \toprule
    \textbf{dataset} & \textbf{est. precision ($\hat \sigma$)} & \textbf{est. recall ($\hat \sigma$)} & \textbf{scope} \\
    \midrule
    Staff 1 & 86\% (5.6\%) & 94\% (2.1\%)  & 1976 -- Dec. 31, 2022 (U.S. granted)\\
    Staff 2 & 88\% (5.6\%) & 95\% (2.0\%) & 1976 -- Dec. 31, 2022 (U.S. granted)\\
    Israeli Benchmark & 79\% (NA) & 94\% (NA) & 1976 -- 1999 (U.S. granted)\\
    \cite{Li2014}'s Benchmark & 91\% (2.7\%) & 91\% (5.0\%) & 1976 -- 2010 (U.S. granted)\\
    \bottomrule
    \end{tabular}
\end{table}

@@

There are two main resulting products for this work. The first is to provide PatentsView data users with an appropriate understanding of PatentsView data quality. Our performance estimates indicate that, despite an overall accurate disambiguation, there is significant room for \ob{improvement. Notably, the current disambiguation over-estimates the number of matching inventor mention pairs.} The second product is the set of tools needed for methodological research and model comparison. Given our evaluation methodology, we can now reliably compare algorithms and decide with confidence on changes which will affect users.

{One important topic for future work is the quantification of uncertainty associated with errors in the hand-disambiguation process. This is challenging problem given the lack of validation information available. Surveying inventors to validate the hand-disambiguation process would be one approach to explore for this. Sensitivity analyses involving potential errors in the hand-disambiguation process could also be informative. We refer the reader to \citep{bailey2017well} for a state-of-the-art hand-labeling study which evaluated human review accuracy.}

{Another important topic for future work is the development of estimators for additional performance metrics. As we have seen in the simulation study, the generically-derived estimators (e.g., from lemma \ref{lemma:cluster_sampling_representation}) do not perform as well as estimators derived using specific properties of pairwise precision and recall (lemma \ref{lemma:block_sampling_representation}). As such, care should be taken to obtain efficient estimators for every metric of interest.}

\ob{Finally, we note that the performance of estimators can degrade when dealing with heavy-tailed cluster size distributions. Bias of ratio estimators can be high in this case, especially when using small samples and when sampling clusters uniformly at random rather than with probability proportional to their size. Model-based estimators which exploit known properties of the cluster size distribution could be developed to improve estimation accuracy in this situation.}

@@

Olivier Binette led the project and wrote most of the manuscript. Sokhna A York and Emma Hickerson carried out data collection by manually reviewing inventor clusters. Youngsoo Baek provided bias adjustment and uncertainty quantification for ratio estimators. Sarvo Madhavan was a technical advisor and contributed to code. Christina Jones was an advisor and project manager. All authors provided input on the manuscript.
